On the surprising effectiveness of a simple matrix exponential derivative approximation, with application to global SARS-CoV-2

### 2023-10-27 manuscript/revision_letter.tex
marcs suggestions and figure numbering fixes in the letter

  That said, we thank you for pointing out your concern.  In response, we have added a new figure (Figure \ref{fig:lowDim}) in the new section ``Empirical Studies'' that helps illustrate the fact that all three algorithms sample from the same distribution.  In order to aid comparison to the exact gradient algorithm (which scales very poorly in state space dimension) we consider a $5$-dimensional state space.  We have also added a second new figure (Figure \ref{fig:truthMean}) to the same section that illustrates how certain Bayesian bridge priors help identify parameters even in small sample settings. Importantly, we use the surrogate-trajectory HMC with the naive derivative in this example.  Finally, we note that the scalable implementation of the affine corrected approximation into BEAST phylogenetic inference software is a matter of ongoing research since it might prevent the efficient outer product coding of the approximate derivative given in Equation 18 of \citet{magee2023random}.
\end{reply}

@@

\setcounter{figure}{4}
\begin{figure}[t]
	\centering
	\includegraphics[width=0.7\linewidth]{figures/randEffects.pdf}
	\caption{Posterior means for exponentiated random effects convey expected multiplicative deviations from the portion of the rate attributable to fixed effects for each corresponding element of the generator matrix.  Notably, we infer a roughly 1.81-fold posterior mean increase in the rate of transitions from the US to Hubei, CN, beyond that portion of the rate which may be explained by fixed effects.  Less pronounced are posterior mean multiplicative increases of 1.34 (US to Italy), 1.27 (US to UK), 1.44 (Netherlands to Italy) and 1.21 (Guangdong, CN, to Hubei, CN).}%\label{fig:randEffs}
\end{figure}

@@

\setcounter{figure}{3}
\begin{figure*}[!t]
	\centering
	\includegraphics[width=0.7\textwidth]{figures/summary2.tree.pdf}
	\vspace{-0.5em}
	\caption{Posterior predictive modes for (unobserved) ancestral
		locations color a phylogenetic tree that describes the
		shared evolutionary history of 285 SARS-CoV-2
		samples.}\label{fig:tree}
\end{figure*}

@@

In short, our original goal was not to prove validity of the general methodology, which we saw as already established through simulation studies, real-world analyses and measure theoretic arguments.

>>>>

  That said, we thank you for pointing out your concern.  In response, we have added a new figure (Figure \ref{fig:lowDim}) in the new section ``Empirical Studies'' that helps illustrate the fact that all three algorithms sample from the same distribution.  In order to aid comparison to the exact gradient algorithm (which scales very poorly in state space dimension) we consider a $5$-dimensional state space.  We have also added a second new figure (Figure \ref{fig:truthMean}) to the same section that illustrates how certain Bayesian bridge priors help identify parameters even in small sample settings. Importantly, we use the surrogate-trajectory HMC with the naive derivative in this example.
\end{reply}

@@

\setcounter{figure}{6}
\begin{figure}[t]
	\centering
	\includegraphics[width=0.7\linewidth]{figures/randEffects.pdf}
	\caption{Posterior means for exponentiated random effects convey expected multiplicative deviations from the portion of the rate attributable to fixed effects for each corresponding element of the generator matrix.  Notably, we infer a roughly 1.81-fold posterior mean increase in the rate of transitions from the US to Hubei, CN, beyond that portion of the rate which may be explained by fixed effects.  Less pronounced are posterior mean multiplicative increases of 1.34 (US to Italy), 1.27 (US to UK), 1.44 (Netherlands to Italy) and 1.21 (Guangdong, CN, to Hubei, CN).}%\label{fig:randEffs}
\end{figure}

@@

\setcounter{figure}{5}
\begin{figure*}[!t]
	\centering
	\includegraphics[width=0.7\textwidth]{figures/summary2.tree.pdf}
	\vspace{-0.5em}
	\caption{Posterior predictive modes for (unobserved) ancestral
		locations color a phylogenetic tree that describes the
		shared evolutionary history of 285 SARS-CoV-2
		samples.}\label{fig:tree}
\end{figure*}

@@

In short, our original goal was not to prove validity of surrogate-trajectory HMC, which we saw as already established through simulation studies, real-world analyses and measure theoretic arguments.
